---
title: "Week9"
format: html
editor: visual
---

# Approach (Week 9)

## 1. Introduction

For this assignment, I use one of the NY Times APIs to collect real data and analyze it. I chose the Most Popular API, which provides information about the most viewed, shared and emailed articles. I first created an app in the New York Times Developer portal and got an API key. Then I used that key to access the API. The goal of this assignment is to practice working with APIs, converting JSON data into a tidy format, and performing a simple analysis.

## 2. Question

The question is:

*Which sections appear most often among the most viewed New York Times articles in the last 7 days?*

## 3. API Selection

I selected the Most Popular API.

Endpoint used:
	•	/viewed/7.json

This endpoint returns the most viewed articles from the last 7 days, including information such as title, section, author, publication date, and URL.

## 4. Request Details

To make the API request, I used the base URL:

https://api.nytimes.com/svc/mostpopular/v2/viewed/7.json

This API requires an API key to authenticate the request. The key is passed as a query parameter named api-key.

When the API key is not included, the API returns an error message. After including the API key, the request returns the article data in JSON format.

## 5. Data Collection Plan

First, I will use the httr package in R to send a request to the Most Popular API.

Then, I will use the jsonlite package to parse the JSON response into an R object.

From the returned JSON, I can already see that the response includes general information such as status and num_results, and that the main article information is stored inside the results field.

## 6. Data Transformation Plan

Next, I will extract the results field from the JSON response and convert it into a tidy data frame.

After that, I will keep only the columns that are most useful for my analysis, such as:
	•	title
	•	section
	•	byline
	•	published date
	•	source
	•	URL

## 7. Data Cleaning Notes

The JSON response includes many fields, including nested fields such as media information, keywords, and other metadata. Since those fields are not necessary for answering my question, I will remove them and keep only the main variables needed for analysis.

This will make the final data frame easier to read and work with.

## 8. Analysis Plan

After cleaning the data, I will count how many times each section appears in the dataset.

This will help me identify which sections are most common among the most viewed New York Times articles in the last 7 days.

## 9. Expected Outcome

At the end, I expect to produce:
	•	a clean and tidy data frame of NYT articles
	•	a summary table showing article counts by section

This will allow me to clearly answer my question.

# Code Base

First, I load the libraries.

- The httr package is to make the API request.
- The jsonlite package is to read and parse the JSON response.
- The tidyverse package is to clean and organize the data.
```{r}
library(httr)
library(jsonlite)
library(tidyverse)

api_key <- Sys.getenv("NYT_API_KEY") # read my API key from an environment variable
```



- I define the API URL for the Most Popular endpoint.
```{r}
url <- "https://api.nytimes.com/svc/mostpopular/v2/viewed/7.json"

#I send the request to the API and include my API key as a query parameter.
response <- GET(url, query = list(`api-key` = api_key))
```
This sends a GET request to the API. The API key is passed in the request so that the New York Times server can authenticate it.



- Next, I convert the API response into text and then parse it from JSON into an R object.
```{r}
data_json <- content(response, as = "text", encoding = "UTF-8")
data_list <- fromJSON(data_json, flatten = TRUE)
```

```{r}
names(data_list)
```



- The main article data is stored inside the results field, so I extract that part next.
```{r}
articles <- as_tibble(data_list$results)
```



- Then, I keep only the columns that are useful for my analysis.
```{r}
nyt_clean <- articles %>%
  select(
    title,
    abstract,
    section,
    byline,
    published_date,
    source,
    url
  )
```



- I count how many times each section appears in the data
```{r}
section_counts <- nyt_clean %>%
  count(section, sort = TRUE)
```



- Finally, I display both the cleaned data frame and the summary table.
```{r}
nyt_clean
section_counts
```

## Data Cleaning Keys

The JSON response included many fields, including nested media information, keywords, and other metadata. For this assignment, I focused only on the results field because it contained the main article data. I then selected only the columns that were relevant to my question and excluded extra nested fields that were not necessary for the analysis.