---
title: "Project 4"
author: "Sinem K Moschos"
format: html
editor: visual
---

## Approach

### Problem Overview

The goal of this project is to classify documents as spam or ham (non-spam) using a labeled dataset. This means building a model that can learn from existing emails and then predict if a new email is spam or not.

### Data Source

For this project, I use the Apache SpamAssassin Public Corpus, which contains real email messages already separated into spam and non-spam categories.

I downloaded two parts of the dataset:

-	easy_ham for normal emails
-	spam for spam emails

The original dataset stores each email as a separate text file.

### Data Preparation

Because the original dataset contains many separate email files, I created a smaller subset for this project by selecting 100 ham emails and 100 spam emails. Then I combined these emails into one CSV file with two columns: one column for the label and one column for the email text.


```{r}
#| eval: false
ham_path <- "/Users/sinemkilicderemoschos/Downloads/easy_ham"
spam_path <- "/Users/sinemkilicderemoschos/Downloads/spam"

ham_files <- list.files(ham_path, full.names = TRUE)
spam_files <- list.files(spam_path, full.names = TRUE)

ham_sample <- ham_files[1:100]
spam_sample <- spam_files[1:100]

ham_texts <- sapply(ham_sample, read_file)
spam_texts <- sapply(spam_sample, read_file)

emails <- data.frame(
  label = c(rep("ham", length(ham_texts)), rep("spam", length(spam_texts))),
  text = c(ham_texts, spam_texts),
  stringsAsFactors = FALSE
)

write.csv(emails, "project4_emails.csv", row.names = FALSE)
```

This step makes the project easier to manage and more reproducible in R.

After that, I clean the text data by:

-	converting text to lowercase
-	removing punctuation
-	removing numbers
-	removing common stopwords
-	removing extra whitespace

### Training and Testing Split

After preparing the data, I will divide it into:

-	a training dataset to build the model
-	a testing dataset to evaluate the model

This allows me to test how well the model performs on unseen documents.

### Model Selection

For classification, I will use a Naive Bayes model. This model is simple and mostly used for text classification problems like spam detection. It works by calculating probabilities of words belonging to spam or ham.

### Evaluation

Finally, I will use the trained model to predict the class of emails in the test dataset. I will compare predicted results with actual labels to evaluate performance. In addition, I can also test the model on a few new email examples outside the training dataset to demonstrate how the model classifies completely unseen documents.


## Code Base

In this section, I load the prepared email dataset from GitHub, clean the text, split the data into training and testing sets, train a Naive Bayes model, and then use the model to classify emails as spam or ham.

```{r message=FALSE, warning=FALSE}
library(readr)
library(tm)
library(e1071)
library(caret)
```

### Read the Dataset

I read the dataset from GitHub. This dataset contains 200 emails and two columns: one for the label and one for the email text.

```{r}
emails <- read_csv(
  "https://github.com/sinemkilicdere/Data607/raw/refs/heads/main/Week11/Project%204/project4_emails.csv",
  show_col_types = FALSE
)

head(emails)
```

### Prepare the Text and Labels

I clean the text encoding and convert the label column into a factor for classification.

```{r}
emails$text <- sapply(emails$text, function(x) {
  x <- enc2utf8(x)
  x <- iconv(x, from = "UTF-8", to = "ASCII", sub = " ")
  ifelse(is.na(x), "", x)
})

emails$label <- as.factor(emails$label)

table(emails$label)
```

### Split the Data into Training and Testing Data

I use 80% of the data for training and 20% for testing.

```{r}
set.seed(123)

train_index <- createDataPartition(emails$label, p = 0.8, list = FALSE)

train_data <- emails[train_index, ]
test_data  <- emails[-train_index, ]

dim(train_data)
dim(test_data)
```

### Create and Clean the Training Corpus

I create text corpora for the training and testing data. Then I clean the text by making it lowercase, removing punctuation, numbers, stopwords, and extra spaces.

```{r message=FALSE, warning=FALSE}
clean_corpus <- function(text_vector) {
  corpus <- VCorpus(VectorSource(text_vector))
  corpus <- tm_map(corpus, content_transformer(tolower))
  corpus <- tm_map(corpus, removePunctuation)
  corpus <- tm_map(corpus, removeNumbers)
  corpus <- tm_map(corpus, removeWords, stopwords("english"))
  corpus <- tm_map(corpus, stripWhitespace)
  return(corpus)
}

train_corpus <- clean_corpus(train_data$text)
test_corpus  <- clean_corpus(test_data$text)
```

### Build the Document Term Matrix

I convert the cleaned text into document-term matrices so the model can work with word frequencies.

```{r message=FALSE, warning=FALSE}
train_dtm <- DocumentTermMatrix(train_corpus)
test_dtm <- DocumentTermMatrix(test_corpus, control = list(dictionary = Terms(train_dtm)))
```

### Convert Word Counts to Yes/No Values

For Naive Bayes, I convert the word counts into binary values showing whether a word appears or not.

```{r}
convert_counts <- function(x) {
  factor(ifelse(x > 0, "Yes", "No"), levels = c("No", "Yes"))
}

train_dtm_binary <- apply(as.matrix(train_dtm), 2, convert_counts)
test_dtm_binary  <- apply(as.matrix(test_dtm), 2, convert_counts)
```

### Train the Naive Bayes Model

I train the Naive Bayes classifier using the training data.

```{r}
classifier <- naiveBayes(train_dtm_binary, train_data$label)
```

### Predict the Test Emails

I use the model to predict the labels for the test emails.

```{r}
predictions <- predict(classifier, test_dtm_binary)

head(predictions)
```

### Evaluate Model Performance

I compare the predicted labels with the real labels using a confusion matrix.

```{r}
confusionMatrix(predictions, test_data$label)
```

### Test the Model on New Email Examples

I test the model on a few completely new email examples to see how it classifies unseen documents.

```{r message=FALSE, warning=FALSE}
new_emails <- c(
  "Congratulations! You have won a free voucher. Click here to claim.",
  "Hi, just checking if we are still meeting tomorrow.",
  "Limited time offer! Get cash fast with no credit.",
)

new_emails <- sapply(new_emails, function(x) {
  x <- enc2utf8(x)
  x <- iconv(x, from = "UTF-8", to = "ASCII", sub = " ")
  ifelse(is.na(x), "", x)
})

new_corpus <- clean_corpus(new_emails)
new_dtm <- DocumentTermMatrix(new_corpus, control = list(dictionary = Terms(train_dtm)))
new_dtm_binary <- apply(as.matrix(new_dtm), 2, convert_counts)

data.frame(
  email = new_emails,
  predicted_label = predict(classifier, new_dtm_binary)
)
```

