---
title: "Week 3A - Global Baseline Estimate"
author: "Sinem K Moschos"
format: html
editor: visual
---

# Week 3A - Approach

## Global Baseline Estimate Recommendation System

### Introduction

For this assignment, I needed to build a very basic recommendation system using movie ratings. Instead of making personalized recommendations, I used something called the Global Baseline Estimate. This method does not try to guess what a specific user likes. It just looks at overall rating patterns and uses simple averages to recommend movies.

### What data I used

I used the movie ratings data we collected earlier. The dataset includes: • User IDs • Movie names • Ratings given by users

Each row shows how one user rated one movie.

### Steps to Approach Problem

I followed the algorithm provided in the spreadsheet and broke the problem into small steps so it would be easier to understand.

First, I calculated the global average rating. This is just the average of all ratings in the dataset. It represents the starting point for every prediction.

Next, I looked at user behavior. Some users usually give high ratings and some are more strict. So I calculated a user bias, which shows how much each user’s average rating is above or below the global average.

Then, I looked at the movies themselves. Some movies are generally liked more than others. So I calculated a movie bias, which shows how much each movie’s average rating is above or below the global average.

### How the Global Baseline Estimate works

After calculating the global mean, user bias, and movie bias, I combined them using this idea:

The predicted rating equals: • the global average rating • plus the user’s bias • plus the movie’s bias

This gives a baseline prediction for how a movie should be rated, without using any personalized recommendation logic.

### Movie Recommendations

Once I had baseline predictions, I averaged the predicted scores for each movie. Then I sorted the movies from highest to lowest predicted rating.

The movies with the highest predicted ratings are the recommended movies according to the Global Baseline Estimate method.

### Why Makes Sense

This approach is useful because: • It is simple and easy to understand • It works even with limited data • It provides a strong baseline to compare with more advanced recommendation systems later

It helped me understand how recommendation systems can work even without personalization.

### Step 1: Load the libraries

First, I loaded the libraries I needed.
I used tidyverse for data manipulation and readxl to read the Excel file.

```{r}
library(tidyverse)
library(readxl)
```

### Step 2: Load the movie ratings data

Next, I loaded the movie ratings spreadsheet into R.
This file contains the user IDs, movie names, and ratings.

```{r}
ratings_raw <- read_excel("MovieRatings.xlsx")
glimpse(ratings_raw)
```

After loading the data, I standardize column names.

```{r}
ratings <- ratings_raw %>%
  pivot_longer(
    cols = -Critic,
    names_to = "movie",
    values_to = "rating"
  ) %>%
  filter(!is.na(rating)) %>%
  rename(user_id = Critic)
```

### Step 3: Calculate the global average rating

Now I calculated the global mean rating.
This is the average of all ratings in the dataset.
This value is important because it is the starting point for the Global Baseline Estimate.

```{r}
global_mean <- mean(ratings$rating)
global_mean
```

### Step 4: Calculate user bias

In this step, I looked at how each user rates movies on average.
Some users usually give higher ratings, and some give lower ratings.
To capture this, I calculated the user bias.
First, I grouped the data by user ID and calculated each user’s average rating.Then I subtracted the global mean from the user’s average.

```{r}
user_bias <- ratings %>%
  group_by(user_id) %>%
  summarise(
    user_avg = mean(rating),
    .groups = "drop"
  ) %>%
  mutate(user_bias = user_avg - global_mean)

user_bias
```

### Step 5: Calculate movie bias

Next, I did the same thing for movies.
Some movies are generally liked more than others.
So I calculated the movie bias by finding each movie’s average rating and comparing it to the global mean.

```{r}
movie_bias <- ratings %>%
  group_by(movie) %>%
  summarise(movie_avg = mean(rating)) %>%
  mutate(movie_bias = movie_avg - global_mean)

movie_bias
```

### Step 6: Calculate the Global Baseline Estimate

Now I combined everything. I joined the user bias and movie bias back into the original ratings data.
Then I applied the Global Baseline Estimate formula.

```{r}
baseline_estimates <- ratings %>%
  left_join(user_bias, by = "user_id") %>%
  left_join(movie_bias, by = "movie") %>%
  mutate(
    baseline_prediction = global_mean + user_bias + movie_bias
  )

baseline_estimates
```

### Step 7: Create movie recommendations

To recommend movies, I averaged the baseline predictions for each movie. Then I sorted the movies from highest to lowest predicted rating.

```{r}
movie_recommendations <- baseline_estimates %>%
  group_by(movie) %>%
  summarise(
    predicted_rating = mean(baseline_prediction)
  ) %>%
  arrange(desc(predicted_rating))

movie_recommendations
```

### Step 8: Top recommended movies

Finally, I selected the top movies based on the highest predicted ratings.

```{r}
top_movies <- movie_recommendations %>%
  slice_head(n = 5)

top_movies
```

